---
title: "Serg's Project Working File"
author: "Sergio Ozoria"
date: "2025-11-26"
output: github_document
---

# 1. Preparing variables

```{r loading package}
library(tidyverse)
```


```{r}
#  serg - cleaning up name using snake format + concatenating missing values 
poke_tidy = read_csv("./Data/pokemon.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names()

# serg - replacing missing type 2 with none
poke_tidy = poke_tidy |> 
  mutate(
    type2 = ifelse(is.na(type2), "None", type2),
    type1 = str_to_title(type1),
    type2 = str_to_title(type2)
  ) |> 
  distinct(name, .keep_all = TRUE)

# serg - cleaning up abilities (could also create variables for each)
poke_tidy = poke_tidy |> 
  mutate(
    abilities = str_remove_all(abilities, "\\[|\\]|'")
  )

# serg - in case there's missing gender 
poke_tidy = poke_tidy |> 
  mutate(
    genderless = ifelse(is.na(percentage_male), TRUE, FALSE)
  )

# serg - categorical var for plotting/modeling
poke_tidy = poke_tidy |> 
  mutate(
    type1 = factor(type1),
    type2 = factor(type2),
    generation = factor(generation),
    is_legendary = as.logical(is_legendary)
  )
```


```{r}
# eman - convert  numeric columns

poke_tidy=
  poke_tidy |> 
  mutate(across(
    .cols = c(starts_with("against_"), "hp", "attack", "defense", "sp_attack", "sp_defense", "speed", "height_m", "weight_kg", "base_total"),
    .fns = as.numeric
  )) 

```

```{r}
# eman - create total stats

poke_tidy =
  poke_tidy |> 
  mutate(total_stats = hp + attack +defense + sp_attack + sp_defense + speed )
```

```{r}
# eman - create attack/defense ratio (useful for visuals later)

poke_tidy = 
  poke_tidy |> 
  mutate(offense_defense_ratio = attack / defense)
```

```{r}
# eman - starting with basic eda

head(poke_tidy)

ggplot(poke_tidy, aes(total_stats))+
  geom_histogram(bins = 30, fill = "skyblue", color = "black")+
  labs(title = "Distribution of Total Stats")
```

```{r}
# eman - average total stats by type

ggplot(poke_tidy, aes(type1, total_stats))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "Total Stats by Primary Type")
```

```{r}
#eman - height vs weight colored by strength

ggplot(poke_tidy, aes(height_m, weight_kg, color = total_stats))+
  geom_point(alpha = 0.6, size =2)+
  scale_color_viridis_c()+
  labs(title = "Height vs Weight (Colored by total stats")
```

```{r}
# eman - hypothesis testing
# test 1: ANOVA - Do types differ in strength?
# Null hypothesis: mean total stats are equal across all types

anova_type = 
  aov(total_stats ~ type1, data = poke_tidy)
summary(anova_type)
```


```{r}
# eman - hypothesis testing
# test 2: t-test - Are legendary Pokemon stronger?
# Null hypothesis:legendary and non-legendary pokemon have the same mean total stats

t.test(total_stats ~ is_legendary, data = poke_tidy)
```

```{r}
# eman - logistic regression (predicting legendary)

legend_model =
  glm (
    is_legendary ~ total_stats + type1+ generation,
    data = poke_tidy, 
    family = "binomial"
  )
summary(legend_model)

poke_tidy$predicted_legendary =
  predict(legend_model, type = "response")

```

```{r}
#eman - saving clean dataset

write_csv(poke_tidy, "Data/pokemon_clean.csv")
```


```{r, warning=FALSE, message=FALSE}

library(randomForest)
library(caret)

#maria - Hypothesis testing

# Can we predict Pokémon type from stats?

poke_type <- poke_tidy |> 
  select(type1, hp, attack, defense, sp_attack, sp_defense, speed) |> 
  mutate(type1 = as.factor(type1))

# 80/20 train-test split
set.seed(123)

train_index <- createDataPartition(poke_type[["type1"]], 
                                   p = 0.8, list = FALSE)

train_set <- poke_type[train_index, ]
test_set  <- poke_type[-train_index, ]

# Train Random Forest
type_model = randomForest(type1 ~ ., data = train_set, ntree = 500)

# Make predictions
predictions = predict(type_model, newdata = test_set)

# Accuracy
accuracy = mean(predictions == test_set[[1]])
baseline = 1/18
improvement = accuracy - baseline

# Results
results = list(
  Accuracy = round(accuracy * 100, 1),
  Baseline = round(100/18, 1),
  Improvement = round((accuracy - (1/18)) * 100, 1)
)

results

# Feature importance
importance_scores = importance(type_model)
varImpPlot(type_model, main = "Feature Importance for Type Prediction")


# Type-specific accuracy
conf_matrix = table(test_set[["type1"]], predictions)

type_accuracy = diag(conf_matrix) / rowSums(conf_matrix)

type_accuracy_df <- data.frame(
  Type = names(type_accuracy),
  Accuracy = round(type_accuracy * 100, 1)
) |> 
  arrange(desc(Accuracy)) |> 
  na.omit()


# Visualization for Prediction Accuracy
ggplot(type_accuracy_df, 
       aes(x = reorder(Type, Accuracy), 
           y = Accuracy)) +
  geom_col(aes(fill = Accuracy > 23.2)) +
  geom_hline(yintercept = 23.2, linetype = "dashed", color = "red") + 
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "gray"),
    labels = c("Below average", "Above average"),
    name = "Type performance"
  ) +
  labs(
    title = "Random Forest: Type-Specific Prediction Accuracy",
    subtitle = "Accuracy (%) for each Pokémon primary type (test set)",
    x = "Pokémon Type",
    y = "Accuracy (%)"
  ) +
  theme_bw()+
theme(legend.position = "bottom")

```

The Random Forest model achieved 23.2% accuracy in predicting Pokémon  primary type from base stats alone—4.1 times better than the 5.6% baseline  expected from random guessing across 18 types. This represents a 17.6 percentage-point improvement and narrowly approaches our 25% stretch target, demonstrating that Pokémon stats do contain meaningful type-specific patterns.

Types with 0% accuracy likely had very small sample sizes in the test set (1-2 Pokémon), making them vulnerable to misclassification. This highlights a limitation of our dataset size for rare type categories.

Surprisingly, Normal type was most predictable (61.9%), possibly because its balanced, middle-range stat profile is distinctive compared to the more specialized distributions of other types.

## HYPOTHESIS TEST CONCLUSION: 
We REJECT the null hypothesis (H₀). The model's performance exceeds baseline, providing evidence that Pokémon primary types differ systematically in their stat distributions.

Variable importance analysis reveals that Special Attack, Attack and Speed are the strongest type predictors. This suggests offensive capabilities and speed tiers are more type-distinctive than defensive stats—consistent with Pokémon battle mechanics where types often specialize in offensive roles (e.g., Fighting types favor high Attack, Psychic types favor Special Attack).

Type-specific accuracy varied substantially. The easiest types to predict were Normal (61.9%), Fighting (40%), and Fire (40%), likely because these types have distinctive stat profiles. In contrast, several types such as Fairy, Ghost, Ground, Ice, Poison, and Rock had 0% accuracy, reflecting high overlap in stat distributions with other types. This explains why the overall model accuracy (~23%) is moderate, many Pokémon share similar stats despite different types.

From a gameplay perspective, base stats provide useful but incomplete information about type identity. Players should consider both stat distributions and type matchups when building balanced teams.







