---
title: "Project Working File"
author: "Sergio Ozoria"
date: "2025-11-26"
output: github_document
---

```{r loading package}
library(tidyverse)
library(scales)
library(viridis)
library(randomForest)
library(caret)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis
```


```{r}
#  serg - cleaning up name using snake format + concatenating missing values 
poke_tidy = read_csv("./Data/pokemon.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names()

# serg - replacing missing type 2 with none
poke_tidy = poke_tidy |> 
  mutate(
    type2 = ifelse(is.na(type2), "None", type2),
    type1 = str_to_title(type1),
    type2 = str_to_title(type2)
  ) |> 
  distinct(name, .keep_all = TRUE)

# serg - cleaning up abilities (could also create variables for each)
poke_tidy = poke_tidy |> 
  mutate(
    abilities = str_remove_all(abilities, "\\[|\\]|'")
  )

# serg - in case there's missing gender 
poke_tidy = poke_tidy |> 
  mutate(
    genderless = ifelse(is.na(percentage_male), TRUE, FALSE)
  )

# serg - categorical var for plotting/modeling
poke_tidy = poke_tidy |> 
  mutate(
    type1 = factor(type1),
    type2 = factor(type2),
    generation = factor(generation),
    is_legendary = as.logical(is_legendary)
  )
```


```{r}
# eman - convert  numeric columns

poke_tidy =
  poke_tidy |> 
  mutate(across(
    .cols = c(starts_with("against_"), "hp", "attack", "defense", "sp_attack", "sp_defense", "speed", "height_m", "weight_kg", "base_total"),
    .fns = as.numeric
  )) 

```

```{r}
# eman - create attack/defense ratio (useful for visuals later)

poke_tidy = 
  poke_tidy |> 
  mutate(offense_defense_ratio = attack / defense)
```

```{r}
# eman - starting with basic eda

head(poke_tidy)

ggplot(poke_tidy, aes(base_total))+
  geom_histogram(bins = 30, fill = "skyblue", color = "black")+
  labs(title = "Distribution of Total Stats")
```

```{r}
# eman - average total stats by type

ggplot(poke_tidy, aes(type1, base_total))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "Total Stats by Primary Type")
```

```{r}
#eman - height vs weight colored by strength

ggplot(poke_tidy, aes(height_m, weight_kg, color = base_total))+
  geom_point(alpha = 0.6, size =2)+
  scale_color_viridis_c()+
  labs(title = "Height vs Weight (Colored by total stats")
```

```{r}
# eman - hypothesis testing
# test 1: ANOVA - Do types differ in strength?
# Null hypothesis: mean total stats are equal across all types

anova_type = 
  aov(base_total ~ type1, data = poke_tidy)
summary(anova_type)
```


```{r}
# eman - hypothesis testing
# test 2: t-test - Are legendary Pokemon stronger?
# Null hypothesis:legendary and non-legendary pokemon have the same mean total stats

t.test(base_total ~ is_legendary, data = poke_tidy)
```

```{r}
# eman - logistic regression (predicting legendary)

legend_model =
  glm (
    is_legendary ~ base_total + type1+ generation,
    data = poke_tidy, 
    family = "binomial"
  )
summary(legend_model)

poke_tidy$predicted_legendary =
  predict(legend_model, type = "response")

```

```{r}
#eman - saving clean dataset

write_csv(poke_tidy, "Data/pokemon_clean.csv")
```


```{r, warning=FALSE, message=FALSE}

#maria - Hypothesis testing

# Can we predict Pokémon type from stats?

poke_type <- poke_tidy |> 
  select(type1, hp, attack, defense, sp_attack, sp_defense, speed) |> 
  mutate(type1 = as.factor(type1))

# 80/20 train-test split
set.seed(123)

train_index <- createDataPartition(poke_type[["type1"]], 
                                   p = 0.8, list = FALSE)

train_set <- poke_type[train_index, ]
test_set  <- poke_type[-train_index, ]

# Train Random Forest
type_model = randomForest(type1 ~ ., data = train_set, ntree = 500)

# Make predictions
predictions = predict(type_model, newdata = test_set)

# Accuracy
accuracy = mean(predictions == test_set[[1]])
baseline = 1/18
improvement = accuracy - baseline

# Results
results = list(
  Accuracy = round(accuracy * 100, 1),
  Baseline = round(100/18, 1),
  Improvement = round((accuracy - (1/18)) * 100, 1)
)

results

# Feature importance
importance_scores = importance(type_model)
varImpPlot(type_model, main = "Feature Importance for Type Prediction")


# Type-specific accuracy
conf_matrix = table(test_set[["type1"]], predictions)

type_accuracy = diag(conf_matrix) / rowSums(conf_matrix)

type_accuracy_df <- data.frame(
  Type = names(type_accuracy),
  Accuracy = round(type_accuracy * 100, 1)
) |> 
  arrange(desc(Accuracy)) |> 
  na.omit()


# Visualization for Prediction Accuracy
ggplot(type_accuracy_df, 
       aes(x = reorder(Type, Accuracy), 
           y = Accuracy)) +
  geom_col(aes(fill = Accuracy > 23.2)) +
  geom_hline(yintercept = 23.2, linetype = "dashed", color = "red") + 
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "gray"),
    labels = c("Below average", "Above average"),
    name = "Type performance"
  ) +
  labs(
    title = "Random Forest: Type-Specific Prediction Accuracy",
    subtitle = "Accuracy (%) for each Pokémon primary type (test set)",
    x = "Pokémon Type",
    y = "Accuracy (%)"
  ) +
  theme_bw()+
theme(legend.position = "bottom")

```

The Random Forest model achieved 23.2% accuracy in predicting Pokémon  primary type from base stats alone—4.1 times better than the 5.6% baseline  expected from random guessing across 18 types. This represents a 17.6 percentage-point improvement and narrowly approaches our 25% stretch target, demonstrating that Pokémon stats do contain meaningful type-specific patterns.

Types with 0% accuracy likely had very small sample sizes in the test set (1-2 Pokémon), making them vulnerable to misclassification. This highlights a limitation of our dataset size for rare type categories.

Surprisingly, Normal type was most predictable (61.9%), possibly because its balanced, middle-range stat profile is distinctive compared to the more specialized distributions of other types.

## HYPOTHESIS TEST CONCLUSION: 
We REJECT the null hypothesis (H₀). The model's performance exceeds baseline, providing evidence that Pokémon primary types differ systematically in their stat distributions.

Variable importance analysis reveals that Special Attack, Attack and Speed are the strongest type predictors. This suggests offensive capabilities and speed tiers are more type-distinctive than defensive stats—consistent with Pokémon battle mechanics where types often specialize in offensive roles (e.g., Fighting types favor high Attack, Psychic types favor Special Attack).

Type-specific accuracy varied substantially. The easiest types to predict were Normal (61.9%), Fighting (40%), and Fire (40%), likely because these types have distinctive stat profiles. In contrast, several types such as Fairy, Ghost, Ground, Ice, Poison, and Rock had 0% accuracy, reflecting high overlap in stat distributions with other types. This explains why the overall model accuracy (~23%) is moderate, many Pokémon share similar stats despite different types.

From a gameplay perspective, base stats provide useful but incomplete information about type identity. Players should consider both stat distributions and type matchups when building balanced teams.


```{r}
# Serg - Hypothesis testing using linear regression
  # RQ: Is there a relationship between capture rate and overall combat effectiveness?

# serg - selecting analytic sample + covariates

library(gtsummary)
library(sjPlot)

capture_stats <- poke_tidy |>
  mutate(
    capture_rate = as.numeric(capture_rate),
    percentage_male = as.numeric(percentage_male),
    
    generation = factor(
      case_when(
        generation %in% c("1","2","3") ~ "Early",
        generation %in% c("4","5","6","7") ~ "Modern"
      ),
      levels = c("Early", "Modern")
    )
  ) |>
  filter(
    is_legendary == FALSE,
    if_all(c(capture_rate, base_total, height_m,
             percentage_male, weight_kg, generation), ~ !is.na(.x))
  )
  
# running linear regression model

capture_stats_lm = lm(
  base_total ~ capture_rate + height_m + 
    weight_kg + percentage_male + generation,
  data = capture_stats
)

summary(capture_stats_lm)

tbl_regression(
  capture_stats_lm,
  intercept = TRUE,
  estimate_fun = ~ style_number(.x, digits = 2),
  pvalue_fun = ~ style_pvalue(.x),
  label = list(
    capture_rate ~ "Capture Rate",
    height_m ~ "Height (m)",
    weight_kg ~ "Weight (kg)",
    percentage_male ~ "Percentage Male",
    generation ~ "Generation (ref = Early)"
  )
) |> 
  modify_header(label = "**Predictors**")
```

## Capture Rate vs Performance Analysis

As Pokémon data nerds, we were interested in examining whether a Pokemon's capture rate is related to its overall combat potential, which is relevant information for casual Pokémon players and competitive trainers. Thus, we examined whether Pokémon that are harder to catch tend to have higher base stat totals. 

Our conceptual hypothesis is that performance contributes to capture difficulty, meaning that stronger Pokémon are designed to be more challenging to catch. To evaluate this relationship, we tested the following hypotheses:

* H₀: There is no linear relationship between `capture_rate` and `base_total` among non-legendary Pokémon, after accounting for key demographic variables

* H₁: There is a significant negative relationship between `capture_rate` and `base_total` (harder-to-catch Pokémon tend to be stronger), after accounting for key demographic variables

#### Measures

* Exposure (predictor):`capture_rate` represents how easy a Pokémon can be caught, meaning that higher values = easier to catch. This variable was cleaned and recoded as a numeric, continuous variable to correctly model the linear relationship and test our hypothesis.

* Outcome (dependent): `base_total` is the sum of a Pokemon's six base total stats, including `hp`, `attack`, `defense`, `sp_attack`, `sp_defense`, and `speed`. This variable will serve as a proxy to assess overall combat effectiveness.  

#### Covariates

Several species-characteristics were included as covariates in the linear regression model to address potential confounding and improve model precision.
These variables were chosen based on their plausible relationships with both `capture_rate` (main predictor) and `base_total` (main outcome of interest), as well as their relevancy to the universe design and any other factors that could bias our results. These covariates included:

* `height_m`: This represents the physical size of the Pokémon. This variable was selected as a covariate because larger species could exhibit higher base stats and may have lower capture rates, making height a potential confounder. 
* `weight_kg`: This represents the body mass index of the Pokémon. This variable was selected as a covariate because heavier Pokémon could have higher combat stats, as well as influence catch rate difficulty.
* `percentage_male`: This represents the proportion of male species in the dataset.Some species may have sex-based differences when it comes to stats, so sex distribution could correlate with combat potential. This variable also helps account for any sex/gender related differences. 
* `generation`: This represents the generation in which the Pokémon was first introduced. This variable was accounted for, as adjusting for it could control for systematic differences across generations due to temporal decision-making around stats or design based on generation.

Altogether, these covariates are meant to adjust for any potential differences across species, reduce biased estimates, and isolate the association between capture rate and overall base stats to assess combat effectiveness.

#### Interpretation of regression analysis

Among all non-legendary Pokémon included our cross-generational dataset of the Pokémon universe, we observed a significant negative association between `capture_rate` and `base_total`. More specifically, for every one-unit increase in `capture_rate`, meaning easier to catch, `base_total` decreased by about 0.87 points **(β = −0.87, 95% CI: −0.96, −0.79; p < 0.001)**. These results suggest that species designed to be harder to catch tend to possess higher combat power than those who are easier to catch.

At the 5% significance level, the association between `capture_rate` and `base_total` is statistically significant, after adjusting for `height_m`, `weight_kg`, `percentage_male`, and `generation`. Therefore, we reject the null hypothesis that there is no linear relationship between the exposure and outcome of interest. 

```{r}

# serg - running two sample t test to explore whether dual-type pokemon have statistical advantages over single-type ones
poke_2tt = poke_tidy |> 
  mutate(
    type2 = str_to_title(type2),
    dual_type = if_else(type2 == "None", "Single", "Dual")
  )

tt_result = t.test(
  base_total ~ dual_type,
  data = poke_2tt,
  var.equal = FALSE
)

tt_result

poke_2tt |> 
  group_by(dual_type) |> 
  summarise(
    n = n(),
    mean_stat = mean(base_total, na.rm = TRUE),
    sd_stat = sd(base_total, na.rm = TRUE)
  )

ggplot(poke_2tt, aes(x = dual_type, y = base_total, fill = dual_type)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  labs(
    x = "Pokémon Typing",
    y = "Total Base Stats",
    title = "Single-Type vs Dual-Type Pokémon: Total Stats Comparison"
  )
```

## A Two-Sample T-Test of Dual-Type vs Single-Type Pokémons

We next examined whether dual-type Pokémon have a statistical advantage over single type Pokémon in terms of overall combat strengthe. More specifically, we wanted to test whether the total base stats differ between single_type and dual-type Pokémons. 

The following analysis will explore these hypotheses:

* H₀: There is no difference in mean `base_total` between single-type and dual-type Pokémon

* H₁: There is a difference in mean `base_total` between single-type and dual-type Pokémon

#### Measures

We first created a binary type variable (i.e., `dual_type`), where we recoded observations as **Single** for Pokémon with only a primary type (i.e., `type2 == "None"`) and **Dual** for those with both primary and secondary types. Our outcome variable, `base_total`, represents the sum of each Pokemon's six base stats, including `hp`, `attack`, `defense`, `sp_attack`, `sp_defense`, and `speed`, which will serve as a proxy to measure overall combat potential.


To compare mean `base_total` between single type and dual type species, we used a two-sample t-test, which allows for unequal variances, with `base_total` as the continuous outcome and `dual_typ` as the grouping variable. We also plotted the distribution of total stats using violin and boxplots to assess overlap and spread between both groups.

#### Results

On average, Dual-type Pokémon had higher total base stats than single-type Pokémon, as evidenced by a mean of 445.9 compared to 409.9. The difference of about 36.5 points higher tell us there results were statistically significant, as noted below:

* t = 4.39
* df = 796.5
* p = 1.30 × 10^5
* 95% CI: 20.2 to 52.9

Given these results and because the confidence interval does not include the value of zero and the p-value is far below 0.05%, we reject the null hypothesis that there are no mean differences between single-type and dual-type Pokémons. These results show that dual-type species tend to have higher base total stats than those with only one type.

#### Visualization Interpretation & Conclusions

To visualize the distribution of total base stats between dual-type and single-type Pokémons, **a violin, box-plot was created**. As noted on the plot, we see that the distribution for dual-type Pokémons shifted upwards, indicating higher median and a denser cluster of higher-stat Pokémon than those with a single-type. Both groups, however, do show variability, with dual-type Pokémon dominating the upper half of the stat range. The overlap between both groups lets us know that type alone does not determine strengthen. However, we can conclude that the average advantage of dual-type Pokémons is consistent.

Overall, although both groups do include strong and weak species, having two types is associated with a statistical advantage in overall combat potential. It is important to consider that generation design differences may play a role in this association, as later generations could emphasize duality in Pokémon species as more powerful, and, by extension, over better stat distribution. We can also conclude that having a second type offers more resistance and flexibility in combat against single-type Pokémons and could serve as advantage in moderating the effect of an effectiveness against one of the two types in dual-type Pokémons.

# Project Report

## 1. Motivation

As lifelong Pokémon players, 

## 2. Initial questions:

## 3. Data:

### 3.1. Data source

Our data source came from []

### 3.2. Data cleaning rocess

### 3.3. Final dataset
