---
title: "Statistical and Machine Learning Analysis"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(caret)
library(randomForest)
theme_set(theme_minimal())
```

```{r data}
poke <- readr::read_csv("Data/pokemon_clean.csv", show_col_types = FALSE) |>
  mutate(
    type1 = as.factor(type1),
    generation = as.factor(generation),
    is_legendary = as.logical(is_legendary)
  )
```

## ANOVA: Do primary types differ in total base stats?

```{r anova}
anova_fit <- aov(base_total ~ type1, data = poke)
anova_results <- broom::tidy(anova_fit)
anova_results
```

## Two-sample t-test: Are legendary Pokémon stronger?

```{r ttest}
ttest_results <- t.test(base_total ~ is_legendary, data = poke)
ttest_results
```

## Logistic regression: Predicting legendary status

```{r logistic}
glm_fit <- glm(
  is_legendary ~ base_total + type1 + generation,
  data = poke,
  family = "binomial"
)
summary(glm_fit)
```

## Random Forest: Predicting primary type from base stats

```{r rf, message=FALSE, warning=FALSE}
set.seed(123)
poke_type <- poke |>
  select(type1, hp, attack, defense, sp_attack, sp_defense, speed) |>
  mutate(type1 = factor(type1))

idx <- createDataPartition(poke_type$type1, p = 0.8, list = FALSE)
train <- poke_type[idx, ]
test  <- poke_type[-idx, ]

rf_fit <- randomForest(type1 ~ ., data = train, ntree = 500)
pred   <- predict(rf_fit, newdata = test)

acc <- mean(pred == test$type1)
baseline <- 1/length(levels(train$type1))

list(
  Accuracy  = round(100*acc, 1),
  Baseline  = round(100*baseline, 1),
  Improvement_pts = round(100*(acc - baseline), 1)
)
```

### Brief interpretation (to customize)

- ANOVA: Inspect the p-value to assess differences in mean total stats across primary types.
- t-test: Typically shows legendaries have higher total stats; confirm via mean difference and confidence interval.
- Logistic regression: `base_total` often strongly predicts `is_legendary`; type and generation may add nuance.
- Random Forest: Expect performance above the naive baseline; examine variable importance with `varImpPlot(rf_fit)` if desired.
