---
title: "About our Trainers"
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
library(htmltools)
knitr::opts_chunk$set(echo = FALSE)
```

<style>
.hero { margin: 12px 0 24px; }
.hero h1 { margin: 0 0 6px; font-family: Montserrat, Arial, sans-serif; }

.trainer-grid {
  display: grid;
  grid-template-columns: 1fr;
  gap: 24px;
}

.trainer-card.trainer-card--horiz {
  display: grid;
  grid-template-columns: minmax(520px, 42%) 1fr; /* large image column on the left */
  align-items: start;
  gap: 22px;
  border: 1px solid #e6e6e6;
  border-radius: 12px;
  padding: 16px;
  background: #fff;
}

.trainer-card.trainer-card--horiz > a { display: block; width: 100%; }

.trainer-card.trainer-card--horiz img {
  display: block;
  width: 100%;
  height: auto;          /* preserve aspect ratio */
  object-fit: contain;   /* show entire image */
  background: #fafafa;
  border-radius: 10px;
}

.trainer-card__body { min-width: 0; }
.trainer-card__name { margin: 0 0 6px; font-family: Montserrat, Arial, sans-serif; }
.trainer-card__body p { margin: 6px 0; }
.trainer-card__body a { color: #1a3fb0; text-decoration: none; }
.trainer-card__body a:hover { text-decoration: underline; }

@media (max-width: 880px) {
  .trainer-card.trainer-card--horiz { grid-template-columns: 1fr; }
}
</style>

```{r trainers}
# IMPORTANT:
# - Avoid spaces and home (~) paths for images when knitting; use relative paths in the project (e.g., "assets/images/serg-team.png").
# - If you must keep spaces, URL-encode them as %20, e.g., "assets/images/serg%20team.png".
# - Recommended: rename files to use hyphens and no spaces.

trainers <- data.frame(
  name   = c(
    "Sergio Ozoria Ramírez (smo2147)",
    "Maria Beg (mb5357)",
    "Eman Ibrahim (ei2291)",
    "Spencer Riddell (swr2118)"
  ),
  github = c("sergozoria", "mb5357", "ei2291", "spencerriddell"),
  avatar = c(
    "/assets/images/serg_team.png",          
    "/assets/images/maria_team.png",
    "/assets/images/eman_team.png",
    "/assets/images/spence_team.png"        
  ),
  bio = c(
    paste0(
      "<strong>Preferred Name:</strong> Serg (he/him)<br>",
      "<strong>Program:</strong> MPH in Population and Family Health<br>",
      "<strong>Certificate:</strong> Public Health Research Methods<br>",
      "<strong>From:</strong> Pastoria City, Sinnoh<br>",
      "<strong>Starter Pokémon:</strong> Chimchar<br><br>",
      "<strong>Background:</strong> Grew up near the Great Marsh, spending childhood studying the behaviors of local wild Pokémon and leveraging data science techniques to gain a combat advantage in Pokémon battles. By putting his Pokémon data science nerd on, he developed a talent for reading opponents’ strategies in battle and a love for statistical modeling. Based on this Pokéstats work, he spends most of his time using data to build perfect individual values (IVs) in key stats for the purpose of competitive play.<br><br>",
      "<strong>Training Style:</strong> Uses typing advantage and a carefully selected move-set rather than brute force to balance out weaknesses and maximize advantage in battle, even when the odds aren’t in his favor. Although Serg likes to maintain a balanced roster, he loves Fire and Fighting-type Pokémon the most.<br><br>",
      "<strong>Overall Goal:</strong> He wants to become a field researcher to document the diverse Pokémon ecosystem across the Sinnoh region and to uncover the secrets of the region’s legendary Pokémon, including Darkrai, Regigigas, and Heatran.<br><br>",
      "<strong>Contributions:</strong> Project Design, Data Sourcing/Identification, Data Cleaning, Hypothesis Testing, Regression Modeling, Data Interpretation, Manuscript/Final Report Development."
    ),
    paste0(
      "<strong>Preferred Name:</strong> Maria (she/her)<br>",
      "<strong>Program:</strong> MPH in Population and Family Health<br>",
      "<strong>Certificate:</strong> Public Health Research Methods<br>",
      "<strong>From:</strong> Jubilife City, Sinnoh<br>",
      "<strong>Starter Pokémon:</strong> Squirtle<br><br>",
      "<strong>Background:</strong> Growing up in Jubilife City, Maria was captivated by the diversity of Pokémon ecosystems around Sinnoh. From observing wild Pokémon in their habitats to studying their behavior, she developed a passion for exploration and discovery. Her curiosity and analytical mind have made her adept at uncovering patterns in both Pokémon behavior and battle strategy.<br><br>",
      "<strong>Training Style:</strong> Maria approaches battles like fieldwork—careful observation, strategic planning, and adaptability are key. She relies on type advantages, well-chosen move sets, and clever tactics, preferring Psychic-type Pokémon for their intelligence and versatility.<br><br>",
      "<strong>Overall Goal:</strong> She aspires to become a Pokémon Explorer, traveling across Sinnoh to document Pokémon habitats, study rare species, and uncover the secrets of legendary Pokémon—all while honing her skills and building a team ready for any challenge.<br><br>",
      "<strong>Contributions:</strong> Project Design, Exploratory Data Analysis, Hypothesis Testing, Data Interpretation, Regression Modeling, Manuscript/Final Report Development."
    ),
    paste0(
      "<strong>Preferred Name:</strong> Eman Ibrahim (she/her)<br>",
      "<strong>Program:</strong> MPH in Epidemiology<br>",
      "<strong>Certificate:</strong> Public Health & Humanitarian Action<br>",
      "<strong>From:</strong> Lumiose City, Kalos<br>",
      "<strong>Starter Pokémon:</strong> Froakie<br><br>",
      "<strong>Background:</strong> Eman grew up surrounded by diverse Pokémon habitats, where she became fascinated not only by patterns in strength and evolution, but also by the stories behind why certain Pokémon thrive in difficult environments. She has always approached problems with a blend of intuition and curiosity—asking deeper questions about why differences exist rather than just observing them. As she moved into epidemiology and humanitarian work, this instinctive way of seeing the world evolved into a love for research, data interpretation, and understanding populations through both numbers and narratives.<br><br>",
      "<strong>Training Style:</strong> Eman’s training style balances analysis with instinct. She relies on data, visual patterns, and statistical reasoning, but also trusts her ability to read context, notice subtle details, and draw meaning from the “story” behind the numbers. Whether in Pokémon battles or in research, she pairs evidence with intuition—making decisions that are both thoughtful and adaptive.<br><br>",
      "<strong>Overall Goal:</strong> Eman hopes to become a Pokémon population researcher who studies how environment, rarity, and external pressures shape strength and development. She aims to explore how different regions create different Pokémon outcomes, combining scientific analysis with real-world storytelling to understand the full picture.<br><br>",
      "<strong>Contributions:</strong> Data Cleaning/Preprocessing, Exploratory Data Analysis/Visualization, Hypothesis Testing, Regression Modeling, Data Interpretation, Manuscript/Final Report Development."
    ),
    paste0(
      "<strong>Preferred Name:</strong> Spence (she/her)<br>",
      "<strong>Program:</strong> MPH in Epidemiology<br>",
      "<strong>Certificate:</strong> Applied Biostatistics and Public Health Data Science<br>",
      "<strong>From:</strong> Castelia City, Unova<br>",
      "<strong>Starter Pokémon:</strong> Oshawott<br><br>",
      "<strong>Background:</strong>. Raised amidst Castelia’s bustling harbor, Spence grew up charting currents, cataloging marine Pokémon behavior, and learning how environments shape health. She blends epidemiologic methods with biostatistics and public health data science to study transmission dynamics, whether that’s a pathogen in a port city or status effects spreading through a battle. Her field notes with Oshawott evolved into full-fledged dashboards tracking encounter rates, type distributions, and move efficacy. With a soft spot for aquatic ecosystems, she’s especially interested in how water-quality metrics and coastal infrastructure impact both human communities and Water-type habitats.<br><br>",
      "<strong>Training Style:</strong> Prefers control and resilience over brute force. Spence uses hazard management (Recover, Wish, and screens), type resistances, and status mitigation to outlast opponents while setting up decisive win conditions. She leans on Water-types like Vaporeon, Jellicent, Mega Starmie, and Spheal for bulk, utility, and reliable coverage, then surprises foes with Cinccino’s multi-hit pressure to break through stalls. Careful team synergy, pivoting, and data-informed move selection help her counter meta threats and turn small advantages into checkmates.<br><br>",
      "<strong>Overall Goal:</strong> To become a field epidemiologist who maps health risks across coastal and riverine regions, integrating Pokémon ecology with public health surveillance. She aims to build open, data-driven tools that protect communities and conserve Water-type habitats.<br><br>",
      "<strong>Contributions:</strong> Project Design, Site Design/Development, Data Analysis, Data Visualization, Data Interpretation."
    )
  ),
  stringsAsFactors = FALSE
)

card_for <- function(name, github, avatar, bio) {
  tags$div(
    class = "trainer-card trainer-card--horiz",
    tags$a(
      href   = paste0("https://github.com/", github),
      target = "_blank", rel = "noopener",
      tags$img(src = avatar, alt = name)
    ),
    tags$div(
      class = "trainer-card__body",
      tags$h3(class = "trainer-card__name", name),
      tags$p(
        tags$a(
          href   = paste0("https://github.com/", github),
          target = "_blank", rel = "noopener",
          paste0("github.com/", github)
        )
      ),
      tags$p(HTML(bio))
    )
  )
}

cards <- lapply(seq_len(nrow(trainers)), function(i) {
  x <- trainers[i, , drop = FALSE]
  card_for(x$name, x$github, x$avatar, x$bio)
})

hero <- tags$div(
  class = "hero",
  tags$h1("Meet the Trainers"),
  tags$p(style = "margin: 0.5rem 0 0;", "The team behind Gotta Graph ’Em All")
)

tagList(
  hero,
  tags$div(class = "trainer-grid", do.call(tagList, cards))
)
```